Spatial variation of the native colon microbiota in healthy adults

### 2017-11-09 submission/manuscript.tex
version sent to collabs

Word count: 3,965/5,000

@@

The human colon is an ecosystem comprised of numerous microenvironments
that select for different microbiota. Concentrations of oxygen, water
and anti-microbial peptides change along the gut axis and influence
which microbiota reside in each location. Microenvironments differ not
only longitudinally along the colon, but also radially from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. The identity of these
specific microbiota and communities are important for understanding the
etiology of complex colon diseases such as Colorectal Cancer (CRC) and
Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
accelerated by perturbations the structure of the gut microbiota (1--3).
The manifestations of these diseases are known to vary based upon the
location in which they occur. For instance, CRC that arises in the
distal (left) colon are of hindgut origin and tend to have large
chromosomal alterations indicative of chromosomal instability (1). In
contrast, CRC arising in the proximal (right) colon tumors are of midgut
origin and tend to be sessile and microsatellite instable (MSI with BRAF
and KRAS mutations) (1). In addition to the environmental gradients
within the colon, the distal and proximal sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively (1,4,5). In IBD patients, disease
occurring in the distal colon extending proximally is usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
anywhere along the GI tract, most commonly in the ileum and the cecum
(2). UC presents as continuous disease with only mucosal involvement,
where as CD has skip lesions and full thickness involvement that may
cause abscesses, strictures and fistulas (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbiota and
their metabolites likely vary as well, and may influence the
heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
long-term complication of IBD, the distribution of microbiota is
important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1,3,6). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). Specific bacteria have also been identified in
fecal samples of patients with varying stages of colon tumorigenesis
(8,9). These species include the oral pathogens \emph{Fusobacterium
nucleatum} and \emph{Porphyromonas asacharolytica}. \emph{F. nucleatum}
has also been found to be elevated in the stool and biopsies of patients
with IBD as compared to healthy controls (10,11). Furthermore, studies
of \emph{F. nucleatum} isolated from mucosal biopsies showed that more
invasive \emph{F. nucleatum} positively correlates with IBD disease
level (10). Like many intestinal pathogens, the bacteria appear to have
a high-impact despite being lowly-abundant in the community (2). The
physiology of these rare taxa may contribute to the colonic disease
state. These studies often examined only shed human stool or the small
intestine, preventing fine-resolution analysis of paired samples from
the proximal and distal sides of the colon. Similarly, comparisons of
on- or off-tumor/lesion bacteria rarely have matched tissue from the
other side of the colon from the same, disease-baring patient, limiting
what conclusions can be drawn about the colonic microbiome overall, let
alone at that specific site. Due to these limitations, the contribution
of the gut microbiota to CRC and IBD disease location in the colon is
largely undefined. Characterizing these communities in healthy
individuals could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial spatial variation of the
colon have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed feces or
on samples collected during colonoscopy or surgery. While the latter
method allows investigators to acquire samples from inside the human
colon, typically these procedures are preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer (12,13). As such, what little information we do have about the
spatial distribution of the microbiota in the proximal and distal colon
is confounded by the bowel preparation procedure.

@@

At a baseline visit, subjects gave consent and were given a home
collection stool kit (Zymo). At least one week prior to the scheduled
colonoscopy, subjects collected whole stool at home and shipped the
samples to a research coordinator on ice. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On the procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Subjects were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and jumbo
biopsy forceps used to collect the luminal contents. Two luminal samples
were collected and the contents immediately deposited into RNAlater
(Fischer) and flash-frozen in liquid nitrogen. The forceps were
withdrawn and new biopsy forceps were used to collect mucosal biopsies
on sections of the colon that were pink and free of stool matter. Three
mucosal biopsies were collected and flash-frozen in RNAlater. These
samples comprised the distal colon samples. The sigmoidoscope was then
withdrawn and a pediatric colonoscope was inserted to reach the proximal
colon. Samples were then collected in the same manner as was done in the
distal colon and the colonoscope withdrawn. All samples were stored at
-80\(^\circ\)C.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (14). Sequences were curated
using the mothur software as described previously (15). The sequences
were assigned a taxonomic classification using a naive Bayesian
classifier trained using a 16S rRNA gene training set from the Ribosomal
Database Project (RDP) (16) and clustered into operational taxonomic
units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
of a mock community revealed the error rate to be 0.018\%. Samples were
rarefied to 4231 sequences per sample in order to reduce uneven sampling
bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
1.39.5 (15). \(\theta\)\textsubscript{YC} distances were calculated to
determine the dissimilarity between two samples. Random Forest
classification models were built using the AUCRF R package using a
leave-one-subject out approach (17). For each model the data was split
into a 19-subject training set and a 1-subject test set. The model was
built and cross-validated using AUCRFcv on the training set. The model
was then tested on the left-out patient. This process was repeated
iteratively for all subjects and results plotted as Reciever Operator
Characteristic curves using the pROC R package (18). Resultant models
were used to identify the OTUs that were most important for classifying
each location. Species-level information for sequences of interest was
obtained by aligning the sequences to the GenBank nucleotide databse
using blastn. The species name was only used if the identity score was
\(\ge\) 99\% over the full-length of the contig and matched a single
reference.

@@

16S rRNA gene sequence reads and experiment metadata are available on
the NCBI Sequence Read Archive (SRA) with accession number XXXX. A
reproducible data analysis pipeline can be found at
\url{https://github.com/SchlossLab/Flynn_LRColon_XXXX_2017}.

@@

Luminal and mucosal samples were collected from the proximal and distal
colon of 20 healthy humans that had not undergone bowel preparation
(Fig. 1). Subjects also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on DNA extracted from each
sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (19). Samples had
varying levels of diversity at each site, irrespective of the individual
(Fig. 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

@@

To determine what factors may have driven the differences seen among the
samples, we compared the community dissimilarity between samples from
all subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to matched samples from the other subjects
(Fig. 3C); this is consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (20--22). Thus
interpersonal variation drove the differences between samples more than
whether the sample came from the proximal or distal side of the colon or
from the lumen or mucosa.

\paragraph{Random Forest classification models identify important
Operational Taxonomic Units (OTUs) on each
side}\label{random-forest-classification-models-identify-important-operational-taxonomic-units-otus-on-each-side}

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We used
10-fold cross validation to build the first model to classify the
luminal versus mucosal samples for the proximal and distal sides,
independently (Fig. 4A). The models performed well when classifying
these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that
were most predictive of each site were identified by their greatest mean
decrease in accuracy when removed from the model. For distinguishing the
proximal lumen and mucosa, OTUs affiliated with the \emph{Bacteriodes},
\emph{Actinomyces}, \emph{Psuedomonas} and \emph{Enterobacteraceae} were
included in the best model (Fig. 5A). The model to differentiate between
the distal lumen and mucosa included OTUs affiliated with the
\emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
\emph{Anaerococcus} (Fig. 5B). These results indicated that there were
fine differences between the different sites of the colon, and that
these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples using 10-fold cross validation. The model
performed best when distinguishing the proximal versus distal mucosa
(Fig. 4B, AUC = 0.850) whereas the model to differentiate between the
proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs
included in the model differentiating the distal and proximal mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} (Fig.
6A). The model that attempted to separate the the proximal and distal
lumen included OTUs affiliated with the \emph{Bacteroides},
\emph{Clostridium IV} and \emph{Oscillibacter} (Fig. 6B). Interestingly,
\emph{Anaerococcus} and \emph{Finegoldia} were distinct between the
mucosa and lumen and also helped to differentiate between the proximal
and distal sides.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors
(23). There were 14 samples from 8 subjects with the \emph{F. nucleatum
subsp. animalis} sequences. Of the samples with the highest relative
abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
were from the proximal mucosa and three from the distal mucosa
(Supplementary Fig. S1A). The second most frequent \emph{Fusobacterium}
sequence was OTU472, which aligned with 99\% identity to \emph{F.
varium}. In addition to \emph{F. nucleatum}, \emph{F. varium} has been
associated with IBD (24). Four subjects harbored \emph{F. varium} and
the samples were split evenly between the proximal and distal mucosa
(Supplementary Fig. S1B). OTU152 was similar to the members of the
\emph{Porphyromonas} genus and the most frequent sequence in that OTU
aligned to \emph{Porphyromonas asacharolytica} (99\% over full length),
another bacterium commonly detected and isolated from colorectal tumors.
OTU152 was only detected on the distal mucosa, and in fact was one of
the OTUs the classification model identified as separating distal and
proximal sides (Supplementary Fig. S1C). Among the 11 distal mucosa
samples that were positive for \emph{P. asacharolytica}, the relative
abundances for this OTU ranged from 0.01\% to 16\%. Thus,
disease-associated OTUs could be found in our sample set of 20 healthy
individuals.

@@

Here we identified bacterial taxa that were specific to the lumen and
mucosa of the proximal and distal sides of the human colon using samples
collected during an unprepared colonoscopy of healthy subjects. We found
that all locations contained a range of phyla relative abundances and a
range of diversity, but that there was a wide variability between
subjects. Pairwise comparisons of each of the sites revealed that the
proximal mucosa and lumen were most similar to each other. Further,
comparison of colonoscopy-collected samples with fecal samples
demonstrated that the distal lumen was most similar to feces. Random
Forest models built using OTU relative abundances from each sample
identified microbiota that were particular to each location of the
colon. Finally, we were able to detect some bacterial OTUs associated
with colonic disease in our healthy cohort. Using unprepped
colonoscopies and machine learning, we have identified bacterial taxa
specific to the healthy proximal and distal human colon.

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, casued by differences in age, sex, diet, but also
spatial ``patchiness'' in the gut microbiome. One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(25). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (26). That
said, across our samples, the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to luminal content in humans (4). Hence, the conclusions
we were able to draw from phyla analysis may have been impacted by
inter-subject patchiness.

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (27). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} along the
colon, but given its anaerobic lifestyle and asacharolytic metabolism,
it is not surprising that it resides in the less-oygen-rich and
protein-rich distal mucosa (4). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreased along the colon to rectum (28). In another study,
\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
mutations, molecular features of proximal CRC (29). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in the present study, the
bacterium was spread across the proximal mucosa, distal lumen and distal
mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
suggests that \emph{Fusobacteria} species are more commonly found on
proximal tumors and in biofilms, indicating that it is not only the
presence of the bacterium but the structure of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
only co-occur on CRC tumors but also to synergize to promote
tumorigenesis in an oral cancer model (30) (31). Further analysis of the
distribution and activities of these pathogens along the colon is needed
to elucidate a mechanism for development of CRC or IBD subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (24).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-\(\alpha\) production than strains from healthy individuals
(10), suggesting the bacteria may increase inflammation in the gut as
well (32). \emph{F. varium} isolated from UC patients caused colonic
ulcers in an experimental mouse model (33). \emph{F. varium} was only
detected in three subjects and two of those samples were isolated form
the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
commonly isolated from UC patient biopsies from the ileum or cecum (34),
suggesting this species may exhibit preference for the different
environmental conditions of these gastrointestinal sites. Further work
will assess how gut environment may select for species which may then
cause localized disease.

Specific comparisons of our findings to previously published studies of
spatial variation are confounded by the use of bowel preparation
methods. A rare report of a matched-colonoscopy study sampled 18
patient's colonic mucosa and luminal contents prior to and after bowel
cleansing (35). This study found that mucosal and luminal samples were
distinguishable prior to bowel cleansing, but that bowel preparation
resulted in an increase in shared OTUs between each site (35). After
seven days, bowel cleansing not only made the samples more difficult to
distinguish, but it also decreased the diversity observed across sites.
Bowel preparation clearly biases the representation of microbiota
recovered from sampling the lumen or mucosa.

@@

\hypertarget{ref-Jalanka2014}{}
12. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
et al. Effects of bowel cleansing on the intestinal microbiota. Gut
{[}Internet{]}. BMJ; 2014;64:1562--8. Available from:
\url{https://doi.org/10.1136/gutjnl-2014-307240}

\hypertarget{ref-Harrell2012}{}
13. Harrell L, Wang Y, Antonopoulos D, Young V, Lichtenstein L, Huang Y,
et al. Standard colonic lavage alters the natural state of
mucosal-associated microbiota in the human colon. Singh SR, editor. PLoS
ONE {[}Internet{]}. Public Library of Science (PLoS); 2012;7:e32545.
Available from: \url{https://doi.org/10.1371/journal.pone.0032545}

>>>>

Word count: 3,953/5,000

@@

The human colon is an ecosystem comprised of numerous microenvironments
that select for different microbiota. Concentrations of oxygen, water,
and anti-microbial peptides change along the gut axis and influence
which microbiota reside in each location. Microenvironments differ not
only longitudinally along the colon, but also radially from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. The identity of these
specific microbiota and communities are important for understanding the
etiology of complex diseases such as Colorectal Cancer (CRC) and
Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
accelerated by perturbations the structure of the gut microbiota (1--3).
The manifestations of these diseases are known to vary based upon the
location in which they occur. For instance, CRC that arises in the
distal (left) colon are of hindgut origin and tend to have large
chromosomal alterations indicative of chromosomal instability (1). In
contrast, CRC arising in the proximal (right) colon tumors are of midgut
origin and tend to be sessile and microsatellite instable (MSI with BRAF
and KRAS mutations) (1). In addition to the environmental gradients
within the colon, the distal and proximal sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively (1,4,5). In IBD patients, disease
occurring in the distal colon extending proximally is usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
anywhere along the GI tract, most commonly in the ileum and the cecum
(2). UC presents as continuous disease with only mucosal involvement,
where as CD has skip lesions and full thickness involvement that may
cause abscesses, strictures and fistulas (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbiota and
their metabolites likely vary as well, and may influence the
heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
long-term complication of IBD, the distribution of microbiota is
important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1,3,6). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). Specific bacteria have also been identified in
fecal samples of patients with varying stages of colon tumorigenesis
(8,9). These species include the oral pathogens \emph{Fusobacterium
nucleatum} and \emph{Porphyromonas asacharolytica}. \emph{F. nucleatum}
has also been found to be elevated in the stool and biopsies of patients
with IBD as compared to healthy controls (10,11). Furthermore, studies
of \emph{F. nucleatum} isolated from mucosal biopsies showed that more
invasive \emph{F. nucleatum} positively correlates with IBD disease
level (10). Like many intestinal pathogens, the bacteria appear to have
a high-impact despite being lowly-abundant in the community (2). The
physiology of these rare taxa may contribute to the colonic disease
state. These studies often examined only shed human stool or the small
intestine, preventing fine-resolution analysis of paired samples from
the proximal and distal sides of the colon. Similarly, comparisons of
on- or off-tumor/lesion bacteria rarely have matched tissue from the
other side of the colon from the same, disease-baring patient, limiting
what conclusions can be drawn about the colonic microbiome overall, let
alone at that specific site (12). Due to these limitations, the
contribution of the gut microbiota to CRC and IBD disease location in
the colon is largely undefined. Characterizing these communities in
healthy individuals could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial spatial variation of the
colon have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed feces or
on samples collected during colonoscopy or surgery (5). While the latter
method allows investigators to acquire samples from inside the human
colon, typically these procedures are preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer (13,14). As such, what little information we do have about the
spatial distribution of the microbiota in the proximal and distal colon
is confounded by the bowel preparation procedure.

@@

At a baseline visit, subjects gave consent and were given a home
collection stool kit (Zymo). At least one week prior to the scheduled
colonoscopy, subjects collected whole stool at home and shipped the
samples to a research coordinator on ice. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On the procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Subjects were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and jumbo
biopsy forceps used to collect the luminal contents. Two luminal samples
were collected and the contents immediately deposited into RNAlater
(Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn
and new biopsy forceps were used to collect mucosal biopsies on sections
of the colon that were pink and free of stool matter. Three mucosal
biopsies were collected and flash-frozen in RNAlater. These samples
comprised the distal colon samples. The sigmoidoscope was then withdrawn
and a pediatric colonoscope was inserted to reach the proximal colon.
Samples were then collected in the same manner as was done in the distal
colon and the colonoscope withdrawn. All samples were stored at
-80\(^\circ\)C.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (15). Sequences were curated
using the mothur software as described previously (16). The sequences
were assigned a taxonomic classification using a naive Bayesian
classifier trained using a 16S rRNA gene training set from the Ribosomal
Database Project (RDP) (17) and clustered into operational taxonomic
units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
of a mock community revealed the error rate to be 0.018\%. Samples were
rarefied to 4231 sequences per sample in order to reduce the effects of
uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
1.39.5 (16). \(\theta\)\textsubscript{YC} distances were calculated to
determine the dissimilarity between two samples. Random Forest
classification models were built using the AUCRF R package using a
leave-one-subject out approach (18). For each model the data were split
into a 19-subject training set and a 1-subject test set. The model was
built and cross-validated using AUCRFcv on the training set. The model
was then tested on the left-out patient. This process was repeated
iteratively for all subjects and results plotted as Reciever Operator
Characteristic curves using the pROC R package (19). Resultant models
were used to identify the OTUs that were most important for classifying
each location. Species-level information for sequences of interest was
obtained by aligning the sequences to the GenBank nucleotide databse
using blastn. The species name was only used if the identity score was
\(\ge\) 99\% over the full-length of the contig and matched a single
reference.

@@

16S rRNA gene sequence reads and experiment metadata are available on
the NCBI Sequence Read Archive (SRA) with accession number XXXX. A
reproducible data analysis pipeline can be found at
\url{https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017}.

@@

Luminal and mucosal samples were collected from the proximal and distal
colon of 20 healthy individuals who had not undergone bowel preparation
(Fig. 1). Subjects also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on DNA extracted from each
sample. As expected, each site was primarily dominated by
\emph{Firmicutes} and \emph{Bacteriodetes} (Fig. 2A) (20). Samples had
varying levels of diversity at each site, irrespective of the individual
(Fig. 2B). For example, the proximal mucosa was more diverse than the
distal for some individuals while the opposite was true for others.
Therefore we could not identify a clear pattern of changes in microbial
diversity along the gut axis.

@@

To determine what factors may have driven the differences seen among the
samples, we compared the community dissimilarity between samples from
all subjects (interpersonal) versus samples from within one subject
(intrapersonal). We found that samples from one individual were far more
similar to each other than to matched samples from the other subjects
(Fig. 3C); this is consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (21--23). Thus
interpersonal variation drove the differences between samples more than
whether the sample came from the proximal or distal side of the colon or
from the lumen or mucosa.

\paragraph{Random Forest classification models identify important OTUs
on each
side}\label{random-forest-classification-models-identify-important-otus-on-each-side}

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We built the
first model to classify the luminal versus mucosal samples for the
proximal and distal sides, independently (Fig. 4A). The models performed
well when classifying these samples (proximal AUC = 0.764, distal AUC =
0.908). The OTUs that were most predictive of each site were identified
by their greatest mean decrease in accuracy when removed from the model.
For distinguishing the proximal lumen and mucosa, OTUs affiliated with
the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
\emph{Enterobacteraceae} were included in the best model (Fig. 5A). The
model to differentiate between the distal lumen and mucosa included OTUs
affiliated with the \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and \emph{Anaerococcus} (Fig. 5B). These results
indicated that there were fine differences between the different sites
of the colon, and that these could be traced to specific OTUs on each
side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples. The model performed best when distinguishing the
proximal versus distal mucosa (Fig. 4B, AUC = 0.850) whereas the model
to differentiate between the proximal versus distal lumen performed
poorly (AUC = 0.580). The OTUs included in the model differentiating the
distal and proximal mucosa included members of the \emph{Porphyromonas},
\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
\emph{Peptoniphilus} (Fig. 6A). The model that attempted to separate the
the proximal and distal lumen included OTUs affiliated with the
\emph{Bacteroides}, \emph{Clostridium IV} and \emph{Oscillibacter} (Fig.
6B). Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were
distinct between the mucosa and lumen and also helped to differentiate
between the proximal and distal sides.

@@

Given that specific bacterial species have been associated with
colorectal cancer and IBD, we probed our sample set for these OTUs.
Among our 100 samples, the most frequent sequence associated with the
\emph{Fusobacterium} genus was OTU179, which aligned via blastn to
\emph{Fusobacterium nucleatum subsp animalis} (100\% over full length).
This is the only species of \emph{Fusobacterium} known to have oncogenic
properties and be found on the surfaces of colorectal cancer tumors
(24). There were 14 samples from 8 subjects with the \emph{F. nucleatum
subsp. animalis} sequences. Of the samples with the highest relative
abundance of \emph{F. nucleatum subsp. animalis}, four of the samples
were from the proximal mucosa and three from the distal mucosa
(Supplementary Fig. S1A). The second most frequent \emph{Fusobacterium}
sequence was OTU472, which aligned with 99\% identity to \emph{F.
varium}. In addition to \emph{F. nucleatum}, \emph{F. varium} has been
associated with IBD (25). Four subjects harbored \emph{F. varium} and
the samples were split evenly between the proximal and distal mucosa
(Supplementary Fig. S1B). OTU152 was similar to the members of the
\emph{Porphyromonas} genus and the most frequent sequence in that OTU
aligned to \emph{Porphyromonas asacharolytica} (99\% over full length),
another bacterium commonly detected and isolated from colorectal tumors.
OTU152 was only detected on the distal mucosa, and in fact was one of
the OTUs the classification model identified as separating distal and
proximal sides (Supplementary Fig. S1C). Among the 11 distal mucosa
samples that were positive for \emph{P. asacharolytica}, the relative
abundances for this OTU ranged from 0.01\% to 16\%. Thus,
disease-associated OTUs could be found in our sample set of 20 healthy
individuals.

@@

We identified bacterial taxa that were specific to the lumen and mucosa
of the proximal and distal sides of the human colon using samples
collected during an unprepared colonoscopy of healthy subjects. We found
that all locations contained a range of phylum relative abundances and a
range of diversity, but that there was a wide variability between
subjects. Pairwise comparisons of each of the sites revealed that the
proximal mucosa and lumen were most similar to each other. Further,
comparison of colonoscopy-collected samples with fecal samples
demonstrated that the distal lumen was most similar to feces. Random
Forest models built using OTU relative abundances from each sample
identified microbiota that were particular to each location of the
colon. Finally, we were able to detect some bacterial OTUs associated
with colonic disease in our healthy cohort. Using unprepped
colonoscopies and machine learning, we have identified bacterial taxa
specific to the healthy proximal and distal human colon.

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, casued by differences in age, sex, diet, but also
spatial ``patchiness'' in the gut microbiome. One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(26). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (27). That
said, across our samples, the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to luminal content in humans (4). Hence, the conclusions
we were able to draw from phyla analysis may have been impacted by
inter-subject patchiness.

@@

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (28). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} along the
colon, but given its anaerobic lifestyle and asacharolytic metabolism,
it is not surprising that it resides in the less-oygen-rich and
protein-rich distal mucosa (4). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreased along the colon to rectum (29). In another study,
\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
mutations, molecular features of proximal CRC (30). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in the present study, the
bacterium was spread across the proximal mucosa, distal lumen and distal
mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
suggests that \emph{Fusobacteria} species are more commonly found on
proximal tumors and in biofilms, indicating that it is not only the
presence of the bacterium but the structure of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
only co-occur on CRC tumors but also to synergize to promote
tumorigenesis in an oral cancer model (31,32). Further analysis of the
distribution and activities of these pathogens along the colon is needed
to elucidate a mechanism for development of CRC or IBD subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (25).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-\(\alpha\) production than strains from healthy individuals
(10), suggesting the bacteria may increase inflammation in the gut as
well (33). \emph{F. varium} isolated from UC patients caused colonic
ulcers in an experimental mouse model (34). \emph{F. varium} was only
detected in three subjects and two of those samples were isolated form
the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
commonly isolated from UC patient biopsies from the ileum or cecum (35),
suggesting this species may exhibit preference for the different
environmental conditions of these gastrointestinal sites. Further work
will assess how gut environment may select for species which may then
cause localized disease.

Specific comparisons of our findings to previously published studies of
spatial variation are confounded by the use of bowel preparation
methods. A rare report of a matched-colonoscopy study sampled 18
patient's colonic mucosa and luminal contents prior to and after bowel
cleansing (36). This study found that mucosal and luminal samples were
distinguishable prior to bowel cleansing, but that bowel preparation
resulted in an increase in shared OTUs between each site (36). After
seven days, bowel cleansing not only made the samples more difficult to
distinguish, but it also decreased the diversity observed across sites.
Bowel preparation clearly biases the representation of microbiota
recovered from sampling the lumen or mucosa.

@@

\hypertarget{ref-Flemer633}{}
12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut {[}Internet{]}. BMJ Publishing Group; 2017;66:633--43.
Available from: \url{http://gut.bmj.com/content/66/4/633}

\hypertarget{ref-Jalanka2014}{}
13. Jalanka J, Salonen A, Salojärvi J, Ritari J, Immonen O, Marciani L,
et al. Effects of bowel cleansing on the intestinal microbiota. Gut
{[}Internet{]}. BMJ; 2014;64:1562--8. Available from:
\url{https://doi.org/10.1136/gutjnl-2014-307240}

\hypertarget{ref-Harrell2012}{}
14. Harrell L, Wang Y, Antonopoulos D, Young V, Lichtenstein L, Huang Y,
et al. Standard colonic lavage alters the natural state of
mucosal-associated microbiota in the human colon. Singh SR, editor. PLoS
ONE {[}Internet{]}. Public Library of Science (PLoS); 2012;7:e32545.
Available from: \url{https://doi.org/10.1371/journal.pone.0032545}
